The evolution of gene expression in the term placenta of viviparous mammals

### 2016-03-25 manuscript/manuscript.Rnw
clean up RLN discussion; remove \paragraph{}

\paragraph{Pre-eclampsia \human Lineage Specific Genes}

@@

\subsubsection{\ac{GH} family}

The \ac{GH} family has expanded in anthropoid primates, resulting in
approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
reliably distinguish transcripts between the two). All show
substantial expression in the \human placenta
(\cref{fig:csh1_evolution_tree}). The expression of the \ac{GH} family
in placenta appears to be primate specific, and likely existed at the
time of the last common ancestor between all extant anthropoids
\cite{Papper.ea2009:Ancientoriginofplacental}. There is some evidence
that increased expression of these genes may be correlated with
overall birth weight of offspring in \human
\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
possible that the extreme expression levels of these genes may be due
to high birth weight. \human placental \ac{GH} genes are involved in
increasing blood flow, replacing maternal pituitary expression of
\ac{GH}, as well as trophoblast invasion
\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
and it is likely that they have the same function in \bonobo.

@@

Relaxin (\textit{RLN}) is a hormone that has multiple effects during
pregnancy including increasing maternal blood flow, widening the pubic
bone and relaxing uterine musculature in preparation for labor. In
\dog, \elephant, \mouse, \carmeli, and \horse there is high expression
of relaxin in the term placenta, with little evidence in other species
examined here (Supplemental \cref{fig:rln2_evolution_tree}). \textit{RLN} is known
to be produced in the placenta of the pregnant \dog and is used in the
diagnosis of canine pregnancy
\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
the luteal ovary and placenta of \pig
\cite{Hisaw.ea1948:Relaxininovaryof}, and the placenta of \horse
\cite{Klonisch.ea1997:Placentallocalizationofrelaxin}. However, the
source of relaxin produced by pregnant \elephant remains unclear
\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in relaxin in
\elephant has been suggested to be a major signal for the initiation
of labor \cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive}.
Here we show that the placenta of elephants show high expression of
relaxin and is likely a major source of \textit{RLN} at term. In
contrast to the results of
\cite{Evans.ea1993:mouserelaxingenenucleotide}, we also see expression
of \textit{RLN} in the placenta of \mouse and \carmeli.

>>>>

% \paragraph{Pre-eclampsia \human Lineage Specific Genes}

@@

\subsubsection{Growth hormone family}

The growth hormone family of genes has expanded in anthropoid
primates, resulting in approximately 5 genes in \human (\textit{GH1},
\textit{GH2}, \textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at
least three in \bonobo (two \textit{GH2} and \textit{GH1}; owing to
the similarity between the two \textit{GH1} in \textit{P. troglodytes}
we cannot reliably distinguish transcripts between the two). All show
substantial expression in the \human placenta
(\cref{fig:csh1_evolution_tree}). The expression of the \ac{GH} family
in placenta appears to be primate specific, and likely existed at the
time of the last common ancestor between all extant anthropoids
\cite{Papper.ea2009:Ancientoriginofplacental}. There is some evidence
that increased expression of these genes may be correlated with
overall birth weight of offspring in \human
\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
possible that the extreme expression levels of these genes may be due
to high birth weight. \human placental \ac{GH} genes are involved in
increasing blood flow, replacing maternal pituitary expression of
\ac{GH}, as well as trophoblast invasion
\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
and it is likely that they have the same function in \bonobo.

@@

Relaxin (\textit{RLN}) is a hormone that has multiple effects during
pregnancy including increasing maternal blood flow, widening the pubic
bone and relaxing uterine musculature in preparation for labor. In
\dog, \elephant, \mouse, \carmeli, and \horse there is high expression
of relaxin in the term placenta, with little evidence in other species
examined here (Supplemental \cref{fig:rln2_evolution_tree}). \textit{RLN} is known
to be produced in the placenta of the pregnant \dog and is used in the
diagnosis of \dog pregnancy
\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
the luteal ovary and placenta of \pig
\cite{Hisaw.ea1948:Relaxininovaryof}, and the placenta of \horse
\cite{Klonisch.ea1997:Placentallocalizationofrelaxin}. However, the
source of relaxin produced by pregnant \elephant remains unclear
\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in \textit{RLN} in
\elephant has been suggested to be a major signal for the initiation
of labor \cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive}.
Here we show that the placenta of elephants show high expression of
\textit{RLN} and is likely a major source of \textit{RLN} at term. In
contrast to the results of
\cite{Evans.ea1993:mouserelaxingenenucleotide}, we also see expression
of \textit{RLN} in the placenta of \mouse and \carmeli.
